A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2015-11-08 imaging_demo/ms.tex
Make manuscript MNRAS-compatible
From 44f7935 (master):
=====================

(*) Rename figures to figure*.eps in git

In ms.tex,

(*) Rename figures to figure*.eps
(*) Use MNRAS style file and epic.bib in the current directory

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Basic setup. Most papers should leave these options alone.
\documentclass[a4paper,fleqn,usenatbib]{../mnras}

@@

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favoring densely packed array layouts with 
large numbers of antennas ($\Nant\gtrsim 1000$). Since the complexity of 
traditional correlators scales as $\mathcal{O}(\Nant^2)$, there will be a steep 
cost for realizing the full imaging potential of these powerful instruments. 
Through our generic and efficient E-field Parallel Imaging Correlator (EPIC), 
we present the first software demonstration of a generalized direct imaging 
algorithm, namely, the Modular Optimal Frequency Fourier (MOFF) imager. It takes 
advantage of the multiplication-convolution theorem of Fourier transforms. Not 
only does it bring down the cost for dense layouts to 
$\mathcal{O}(\Nant\log_2\Nant)$ but can also image from irregularly arranged 
heterogeneous antenna. EPIC is highly modular and parallelizable, implemented in 
object oriented Python, and publicly available. We have verified the images 
produced to be equivalent to those produced using traditional techniques to 
within a precision determined by coarseness of gridding. We have also validated 
our implementation on data observed with the Long Wavelength Array (LWA). Antenna 
layouts with a dense filling factor consisting of a large number of antennas such 
as LWA, the Square Kilometre Array, Hydrogen Epoch of Reionization Array, and 
Canadian Hydrogen Intensity Mapping Experiment will gain significant 
computational advantage by deploying EPIC. Inherent availability of calibrated 
time-domain images on digitizer writeout timescales and vastly lower I/O 
bandwidth relative to visibility based systems will make it a prime candidate 
for transient searches of Fast Radio Bursts (FRB) as well as planetary and 
exoplanetary phenomena. 
\end{abstract}

@@

Direct imaging correlators have begun to be explored on deployed arrays including
the Basic Element for SKA Training II (BEST-2) array \citep{fos14}, the Omniscope
\citep{zhe14}, and an earlier incarnation at higher frequencies with the intent
of pulsar timing \citep{oto94, dai00}. However, each of these examples make
assumptions about the redundancy of the array layout, and require the collecting
elements are identical. On the other hand, the MOFF algorithm achieves the same
$\Ngrid \log_2 \Ngrid$ computational scaling without placing any restriction on
antenna placement, can accommodate non-identical beam patterns, and is provably 
optimal \citep{mor11}. This algorithm uses the antenna beam patterns to 
grid the electric field measurements to a regular grid in the software 
holography/A-transpose fashion \citep{mor09,bha08,teg97b} before performing the
spatial FFT. This process has been shown to theoretically produce a data product
identical to images produced from traditional visibility based techniques.

Here we present the first software implementation of the MOFF correlator, and
announce the public release of the E-field Parallel Imaging Correlator (EPIC)
code. EPIC is a highly parallel, object oriented Python package that primarily 
implements the MOFF imaging algorithm besides emulating real-life telescopes and 
FX/XF correlators in software, and includes a visibility based imaging technique 
for reference. It is intended to provide a development platform to test different 
imaging approaches, characterize scaling relations and serve as a stepping stone 
for real-life GPU/FPGA based implementation on telescopes.

We begin with a technical description of the algorithm in \S\ref{sec:math}, 
then discuss our particular implementation in \S\ref{sec:software}. We then 
verify the output data quality from our code in \S\ref{sec:verify} by presenting 
simulated images from both the EPIC correlator and comparing to a simulated FX 
correlator. We also demonstrate the performance with real-world data from the 
LWA. In \S \ref{sec:analysis}, we explore the scalability of the algorithm in the 
context of several array design choices. We identify specific array design 
classes where the EPIC correlator -- is computationally more efficient; and in 
the field of transients, demands significantly lesser I/O bandwidth relative to 
visibility based approaches. We conclude and discuss future research prospects 
in \S\ref{sec:conclusions}.

@@

There are some important differences between the two techniques:
\begin{enumerate}
\item The time-averaging cannot be performed on a stochastic measurement but
  only on its statistical properties. In visibility based imaging, the 
  visibilities measured between antenna pairs represent spatial correlations 
  which can be time-averaged followed by gridding and imaging. However, in MOFF 
  imaging both antenna and gridded electric fields are stochastic and therefore 
  must be imaged and squared before time-averaging. 
\item In visibility based imaging, electric fields measured by antennas are not 
  correlated with themselves and hence lack zero spacing measurements. In 
  contrast, in MOFF imaging, since the gridded electric fields are imaged and 
  squared, they retain information from auto-correlated electric fields at zero 
  spacing and thus yield the true total power of the imaged field.
\end{enumerate}

@@

We have implemented the MOFF imaging technique in our ``E-field Parallel Imaging
Correlator'' -- a highly parallelized Object Oriented Python 
package,\footnote{The E-field Parallel Imaging Correlator (EPIC) package can be 
accessed at https://github.com/nithyanandan/EPIC} now publicly available. Besides 
implementing the MOFF imaging algorithm it also includes visibility based imaging 
using the software holography technique and a simulator for generating electric 
fields from a sky model.

@@

Fig.~\ref{fig:MOFF-flowchart} shows the flowchart for MOFF imaging. The
propagated electric fields are shown on the left at different time stamps,
$t_1\ldots t_\textrm{M}$. At each time stamp, the electric fields measured by
antennas are denoted by $E_1(t)\ldots E_\textrm{N}(t)$. The F-engine performs a
temporal Fourier transform on the electric field time-series to obtain electric
field spectra $\widetilde{E}_1(f)\ldots \widetilde{E}_\textrm{N}(f)$ 
($\widetilde{\mathbf{E}}(\mathbf{a})$ in matrix notation) for each of the 
antennas. Each of the complex antenna gains are calibrated to correct the 
corresponding electric field spectra. These calibrated electric fields are 
gridded using an antenna-based gridding convolution function after which it is 
spatially Fourier transformed and squared to obtain an image cube for every time 
stamp. These images are then time-averaged to obtain the accumulated image 
$I(f)$ ($\mathbf{I}(\hat{\mathbf{s}})$ in matrix notation).
\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_flowchart}
  % \includegraphics[width=\columnwidth]{MOFF_flowchart.eps}
  \caption{A flowchart of MOFF imaging in EPIC. The propagated electric fields
    shown on the left are measured as time-series $E_1(t)\ldots E_\textrm{N}(t)$
    by the antennas $\textrm{A}_1\ldots \textrm{A}_\textrm{N}$ which are then 
    Fourier transformed by the F-engine to produce electric field spectra 
    $\widetilde{E}_1(f)\ldots \widetilde{E}_\textrm{N}(f)$. They are calibrated 
    and gridded. The gridded electric fields $\widetilde{E}_\textrm{g}(f)$ from 
    each time series are imaged to produce images 
    $I_1(f)\ldots I_\textrm{M}(f)$. These images are time-averaged to obtain the 
    final image $I(f)$.}
  \label{fig:MOFF-flowchart}
\end{figure}

Fig.~\ref{fig:FX-flowchart} shows the flowchart for a visibility based software 
holographic imaging from a FX correlator. The antenna-based F-engine is identical 
to that in the MOFF processing. The electric field spectra from each antenna are 
then cross-multiplied in the X-engine with those from all other antennas to 
obtain the visibilities $\widetilde{V}_\textrm{p}(f)$ ($\mathbf{m}(\mathbf{v})$ 
in matrix notation). They are calibrated and time-averaged to obtain 
$\langle \widetilde{V}_\textrm{p}(f)\rangle$ which are then gridded and imaged to 
obtain the image $I(f)$. The $I(f)$ obtained from both techniques are 
theoretically identical as explained in \S\ref{sec:math}.
\begin{figure}
  \includegraphics[width=\columnwidth]{FX_flowchart}
  % \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of visibility based software holographic imaging in EPIC. 
    The FX process flow shares the F-engine with the MOFF process. Following the 
    F-engine, the electric fields pass through the X-engine to produce 
    visibilities $\widetilde{V}_\textrm{p}(f)$ which are calibrated and 
    time-averaged. Then they are gridded to obtain the gridded visibilities 
    $\widetilde{V}_\textrm{g}(f)$ which are then Fourier transformed to obtain 
    the image, $I(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{a2g_mapping}
  % \includegraphics[width=\columnwidth]{a2g_mapping.eps}
  \caption{Block diagram of an antenna-to-grid mapping. A sparse block-diagonal
    matrix of total size $\Ngrid\times \Nant$ is created where each
    block contains roughly the number of pixels covered by the respective kernel.
    The antenna aperture illumination kernels do not have to be identical to each
    other. A discrete set of arbitrarily placed antennas are now placed onto a
    regular grid.}
  \label{fig:a2g-mapping}
\end{figure}

\par\medskip
\noindent {\bf Temporal Fourier transform}
\par\medskip
\noindent This module is common to the MOFF and visibility based imaging 
techniques. Time samples of electric fields measured by the antenna and 
digitized by the A/D converter is Fourier transformed to generate electric 
field spectra. This step can be parallelized across antennas as shown in 
Fig.~\ref{fig:f-engine}. The output is then fed to either MOFF or visibility 
based imaging pipelines.

\begin{figure}
  \includegraphics[width=\columnwidth]{F-engine}
  % \includegraphics[width=\columnwidth]{F-engine.eps}
  \caption{Block diagram of a F-engine. The electric field data streams from
    antennas are Fourier transformed in parallel to generate electric field
    spectra.}
  \label{fig:f-engine}
\end{figure}

@@

In a companion paper to appear soon, we demonstrate a novel calibration 
technique (EPICal) which leverages the data products formed by direct imaging 
correlators to estimate antenna complex gains. This method correlates the antenna 
electric field signals with an image pixel from the output of the correlator in 
the feedback calibration fashion outlined in \citealt{mor11} (illustrated in Fig. 
\ref{fig:MOFF-flowchart} by the arrow leading from the imager to the calibration 
block). Furthermore it allows for arbitrarily complex sky models, and following 
the MOFF algorithm places no restriction on array layout, and accounts for 
non-identical antenna beam patterns. Direction dependent calibration can be 
achieved by correlating antenna signals with output pixels in the direction of 
$N_\textrm{c}$ calibration sources, then fitting for a functional model of the 
sky. Since antennas are only correlated with calibrator pixels, the 
computational complexity scales as $\sim N_\textrm{a} N_\textrm{c}$. % And since 
% only a single correlation is needed for each antenna, the computation complexity 
% scales only as $\sim N_{\mathrm{a}}N_\textrm{c}$, where $N_\textrm{c}$ is the 
% number of calibrators in the field.

@@

In MOFF imaging, these are spatially Fourier transformed followed by a squaring
operation at every time stamp for every frequency channel. In visibility based 
imaging, the spatial Fourier transform is performed only once per integration 
timescale and does not include a squaring operation.

@@

In contrast, visibilities measured by an antenna are statistically stable within
an integration time interval. Hence, they are averaged after calibration as shown
in equation~\ref{eqn:cc-vis}. It is advantageous to average them in visibilities
before imaging because the repeated cost of spatial FFT can be avoided. Since 
this averaging has been performed already on the visibilities over an integration 
timescale, the imaging step has to be performed only once per integration cycle.

@@

In order to verify the accuracy of the EPIC code, we characterize the images 
produced through simulations. We simulate electric field streams from a model 
sky and process the data through both the MOFF and a visibility based imaging 
algorithm. We then compare the output images to demonstrate their equivalence.

@@

We use the EPIC simulator to generate electric field samples from a sky model. In
our simulations, we use 64 frequency channels each of width $\delta f = 40$~kHz, 
10 point sources of flux densities 10~Jy at random locations. The number of
time stamps integrated in one integration cycle was kept at eight where each A/D 
timeseries is $1/\delta f=25\,\mu$s long. We use the MWA array layout 
\citep{bea12} for demonstration. Only the inner 51 tiles within a square bounding
box of 150~m on each side were used. We assumed all tiles are identical and have 
a square shaped electric field illumination footprint 4.4~m on each side.

\subsection{Dealing with antenna auto-correlations}\label{sec:rm-autocorr}

Before the outputs can be compared, we describe the elimination of a minor 
mathematical difference between the two techniques. The squaring operation under 
MOFF imaging in the image plane introduces antenna auto-correlations around the 
zero spacing in the $uv$-plane which are absent in traditional visibility based 
imaging. In order to facilitate a robust comparison between MOFF and visibility 
based imaging techniques, these auto-correlations are removed from the MOFF 
algorithm output, which is otherwise not an essential part of the core algorithm. 
We describe below how they are removed.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{autocorr_uvwts_pbeam}
  % \includegraphics[width=\columnwidth]{autocorr_uvwts_pbeam.png}
  \caption{The auto-correlation of weights of a square shaped antenna aperture
    in the $uv$ plane (left) and the corresponding directional antenna power 
    response on the sky (right) in coordinates specified by direction cosines. 
    The antenna auto-correlation weights are normalized to a sum
    of unity yielding a peak response of unity in the antenna's directional
    power pattern on the sky. The color scale for the directional power 
    pattern is logarithmic. The black circle indicates the sky horizon and
    values beyond it are not physical and hence ignored.}
  \label{fig:autocorr_wts_PB}
\end{figure}

We inverse Fourier transform the squared images and beams back to the $uv$ 
plane and subtract the estimated auto-correlation kernel scaled to the peak 
value centered at the zero spacing pixel. The final averaged image is obtained 
by Fourier transforming the $uv$ plane data and weights with the 
auto-correlations subtracted to the image plane. These images are now comparable 
to those obtained from visibility based imaging. This step of removing 
auto-correlations is optional and required to be performed only once per 
integration timescale and does not add significant cost to the full operation.

@@

Fig.~\ref{fig:MOFF-FX-image} shows the dirty images (top) and synthesized beams
(bottom) obtained with antenna-based MOFF and FX visibility based imaging 
algorithms packaged in EPIC. The antenna auto-correlations that correspond to 
zero spacing have been removed from the correlated weights and data in the 
$uv$ plane, the MOFF image and the corresponding synthesized beam as 
described in \S\ref{sec:rm-autocorr}. The sky positions of the simulated sources 
are indicated by solid black circles. The reconstructed sky image has the 
simulated sources at the expected sky positions in either case. Both algorithms 
result in images and synthesized beams that are well matched with each other. 
Their fluxes are modulated by a multiplicative power pattern corresponding to 
that of a uniform square aperture.

\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_FX_comparison_10_random_source_positions_4_iterations_test_aperture_zoomed}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_10_random_source_positions_4_iterations_test_aperture_zoomed.eps}
  \caption{Dirty images (top) and synthesized beams (bottom) obtained from 
    simulated data using EPIC implementation of antenna-based MOFF algorithm 
    (left) and visibility-based imaging (right). The solid black circles in the
    top panels indicate the simulated source positions. The antenna 
    auto-correlations at zero-spacing have been removed from the MOFF images. 
    The images in either case reconstruct the sources at the right locations 
    with the fluxes expected after multiplication by the antenna power pattern. 
    The synthesized beams from the two algorithms are well matched in size and
    shape. The overall modulation by the power pattern is seen clearly in both
    images.}
  \label{fig:MOFF-FX-image}
\end{figure}

Now we investigate the gridded cross-correlation weights in the $uv$-plane. 
Fig.~\ref{fig:MOFF-FX-uvwts} shows the correlated weights obtained with MOFF 
imaging (left) and visibility based imaging (right). There are two notable 
differences. The first is in the weights around zero spacing. Though both show a
dominant void around zero-spacing, the void obtained with MOFF algorithm shows 
many pixels with non-zero weights. In contrast, the zero-spacing void from 
traditional imaging consists of predominantly zero-valued pixels. The second 
difference is in the weights outside the zero-spacing void. In some regions they 
are different from each other at the few percent level though the sum of weights 
in these regions is the same. For instance, note the difference in weights at 
$(u,v)\approx (2,-10)$. The reason for these differences is described below.

The gridding step in MOFF imaging samples the antenna footprint (either in 
analytic or lookup table formats) at the grid locations. Coverage of grid pixels 
by an antenna footprint may be $\sim 1$ pixel narrower particularly at the 
edge of the footprint along one or both directions relative to that from another 
identical antenna but with a fractionally different location relative to the 
grid. This depends on the exact location of the center of the antenna relative 
to the grid and the coarseness of grid spacing. This first order loss of 
precision of the sampled footprint propagates to second order ($\sim 2$ pixels) 
upon correlation of the discretized weights. In other words, the correlated 
weights may suffer further loss of precision in their sampled footprint after 
correlation of two footprints each of which could be less precise to first 
order. On the other hand, in visibility based imaging, a directly sampled $uv$ 
plane antenna power response (which is theoretically identical to the correlation 
of individual antenna footprints) centered on a baseline has a loss in precision 
at most to first order. Thus, although in the limit of infinitesimally small 
grid spacing they should be identical, the coarseness of grid spacing introduces 
subtle differences between the two.

These differences which are dependent on the coarseness of grid spacing can be 
mitigated by making the grid spacing finer at the expense of increased 
computational cost. Residuals centered around zero spacing can also be lowered 
by subtracting each auto-correlation of antenna weights separately by using the 
shape and extent of the sampled footprint appropriate for that specific antenna 
aperture. This is a general solution applicable even in case of heterogeneous 
antenna arrays and is under active development for EPIC.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Basic setup. Most papers should leave these options alone.
% \documentclass[a4paper,fleqn,usenatbib]{../mnras}
\documentclass[a4paper,fleqn,usenatbib]{mnras}

@@

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favouring densely packed array layouts with 
large numbers of antennas ($\Nant\gtrsim 1000$). Since the complexity of 
traditional correlators scales as $\mathcal{O}(\Nant^2)$, there will be a steep 
cost for realizing the full imaging potential of these powerful instruments. 
Through our generic and efficient E-field Parallel Imaging Correlator (EPIC), 
we present the first software demonstration of a generalized direct imaging 
algorithm, namely, the Modular Optimal Frequency Fourier (MOFF) imager. It takes 
advantage of the multiplication-convolution theorem of Fourier transforms. Not 
only does it bring down the cost for dense layouts to 
$\mathcal{O}(\Nant\log_2\Nant)$ but can also image from irregularly arranged 
heterogeneous antenna. EPIC is highly modular and parallelizable, implemented in 
object oriented Python, and publicly available. We have verified the images 
produced to be equivalent to those produced using traditional techniques to 
within a precision determined by coarseness of gridding. We have also validated 
our implementation on data observed with the Long Wavelength Array (LWA). Antenna 
layouts with a dense filling factor consisting of a large number of antennas such 
as LWA, the Square Kilometre Array, Hydrogen Epoch of Reionization Array, and 
Canadian Hydrogen Intensity Mapping Experiment will gain significant 
computational advantage by deploying EPIC. Inherent availability of calibrated 
time-domain images on digitizer writeout time-scales and vastly lower I/O 
bandwidth relative to visibility-based systems will make it a prime candidate 
for transient searches of Fast Radio Bursts (FRB) as well as planetary and 
exoplanetary phenomena. 
\end{abstract}

@@

Direct imaging correlators have begun to be explored on deployed arrays including
the Basic Element for SKA Training II (BEST-2) array \citep{fos14}, the Omniscope
\citep{zhe14}, and an earlier incarnation at higher frequencies with the intent
of pulsar timing \citep{oto94, dai00}. However, each of these examples make
assumptions about the redundancy of the array layout, and require the collecting
elements are identical. On the other hand, the MOFF algorithm achieves the same
$\Ngrid \log_2 \Ngrid$ computational scaling without placing any restriction on
antenna placement, can accommodate non-identical beam patterns, and is provably 
optimal \citep{mor11}. This algorithm uses the antenna beam patterns to 
grid the electric field measurements to a regular grid in the software 
holography/A-transpose fashion \citep{mor09,bha08,teg97b} before performing the
spatial FFT. This process has been shown to theoretically produce a data product
identical to images produced from traditional visibility-based techniques.

Here we present the first software implementation of the MOFF correlator, and
announce the public release of the E-field Parallel Imaging Correlator (EPIC)
code. EPIC is a highly parallel, object oriented Python package that primarily 
implements the MOFF imaging algorithm besides emulating real-life telescopes and 
FX/XF correlators in software, and includes a visibility-based imaging technique 
for reference. It is intended to provide a development platform to test different 
imaging approaches, characterize scaling relations and serve as a stepping stone 
for real-life GPU/FPGA-based implementation on telescopes.

We begin with a technical description of the algorithm in \S\ref{sec:math}, 
then discuss our particular implementation in \S\ref{sec:software}. We then 
verify the output data quality from our code in \S\ref{sec:verify} by presenting 
simulated images from both the EPIC correlator and comparing to a simulated FX 
correlator. We also demonstrate the performance with real-world data from the 
LWA. In \S \ref{sec:analysis}, we explore the scalability of the algorithm in the 
context of several array design choices. We identify specific array design 
classes where the EPIC correlator -- is computationally more efficient; and in 
the field of transients, demands significantly lesser I/O bandwidth relative to 
visibility-based approaches. We conclude and discuss future research prospects 
in \S\ref{sec:conclusions}.

@@

There are some important differences between the two techniques:
\begin{enumerate}
\item The time-averaging cannot be performed on a stochastic measurement but
  only on its statistical properties. In visibility-based imaging, the 
  visibilities measured between antenna pairs represent spatial correlations 
  which can be time-averaged followed by gridding and imaging. However, in MOFF 
  imaging both antenna and gridded electric fields are stochastic and therefore 
  must be imaged and squared before time-averaging. 
\item In visibility-based imaging, electric fields measured by antennas are not 
  correlated with themselves and hence lack zero spacing measurements. In 
  contrast, in MOFF imaging, since the gridded electric fields are imaged and 
  squared, they retain information from auto-correlated electric fields at zero 
  spacing and thus yield the true total power of the imaged field.
\end{enumerate}

@@

We have implemented the MOFF imaging technique in our ``E-field Parallel Imaging
Correlator'' -- a highly parallelized Object Oriented Python 
package,\footnote{The E-field Parallel Imaging Correlator (EPIC) package can be 
accessed at https://github.com/nithyanandan/EPIC} now publicly available. Besides 
implementing the MOFF imaging algorithm it also includes visibility-based imaging 
using the software holography technique and a simulator for generating electric 
fields from a sky model.

@@

Fig.~\ref{fig:MOFF-flowchart} shows the flowchart for MOFF imaging. The
propagated electric fields are shown on the left at different time stamps,
$t_1\ldots t_\textrm{M}$. At each time stamp, the electric fields measured by
antennas are denoted by $E_1(t)\ldots E_\textrm{N}(t)$. The F-engine performs a
temporal Fourier transform on the electric field time-series to obtain electric
field spectra $\widetilde{E}_1(f)\ldots \widetilde{E}_\textrm{N}(f)$ 
($\widetilde{\mathbf{E}}(\mathbf{a})$ in matrix notation) for each of the 
antennas. Each of the complex antenna gains are calibrated to correct the 
corresponding electric field spectra. These calibrated electric fields are 
gridded using an antenna-based gridding convolution function after which it is 
spatially Fourier transformed and squared to obtain an image cube for every time 
stamp. These images are then time-averaged to obtain the accumulated image 
$I(f)$ ($\mathbf{I}(\hat{\mathbf{s}})$ in matrix notation).
\begin{figure}
  \includegraphics[width=\columnwidth]{figure1}
  % \includegraphics[width=\columnwidth]{MOFF_flowchart}
  % \includegraphics[width=\columnwidth]{MOFF_flowchart.eps}
  \caption{A flowchart of MOFF imaging in EPIC. The propagated electric fields
    shown on the left are measured as time-series $E_1(t)\ldots E_\textrm{N}(t)$
    by the antennas $\textrm{A}_1\ldots \textrm{A}_\textrm{N}$ which are then 
    Fourier transformed by the F-engine to produce electric field spectra 
    $\widetilde{E}_1(f)\ldots \widetilde{E}_\textrm{N}(f)$. They are calibrated 
    and gridded. The gridded electric fields $\widetilde{E}_\textrm{g}(f)$ from 
    each time series are imaged to produce images 
    $I_1(f)\ldots I_\textrm{M}(f)$. These images are time-averaged to obtain the 
    final image $I(f)$.}
  \label{fig:MOFF-flowchart}
\end{figure}

Fig.~\ref{fig:FX-flowchart} shows the flowchart for a visibility-based software 
holographic imaging from a FX correlator. The antenna-based F-engine is identical 
to that in the MOFF processing. The electric field spectra from each antenna are 
then cross-multiplied in the X-engine with those from all other antennas to 
obtain the visibilities $\widetilde{V}_\textrm{p}(f)$ ($\mathbf{m}(\mathbf{v})$ 
in matrix notation). They are calibrated and time-averaged to obtain 
$\langle \widetilde{V}_\textrm{p}(f)\rangle$ which are then gridded and imaged to 
obtain the image $I(f)$. The $I(f)$ obtained from both techniques are 
theoretically identical as explained in \S\ref{sec:math}.
\begin{figure}
  \includegraphics[width=\columnwidth]{figure2}
  % \includegraphics[width=\columnwidth]{FX_flowchart}
  % \includegraphics[width=\columnwidth]{FX_flowchart.eps}
  \caption{A flowchart of visibility-based software holographic imaging in EPIC. 
    The FX process flow shares the F-engine with the MOFF process. Following the 
    F-engine, the electric fields pass through the X-engine to produce 
    visibilities $\widetilde{V}_\textrm{p}(f)$ which are calibrated and 
    time-averaged. Then they are gridded to obtain the gridded visibilities 
    $\widetilde{V}_\textrm{g}(f)$ which are then Fourier transformed to obtain 
    the image, $I(f)$.}
  \label{fig:FX-flowchart}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figure3}
  % \includegraphics[width=\columnwidth]{a2g_mapping}
  % \includegraphics[width=\columnwidth]{a2g_mapping.eps}
  \caption{Block diagram of an antenna-to-grid mapping. A sparse block-diagonal
    matrix of total size $\Ngrid\times \Nant$ is created where each
    block contains roughly the number of pixels covered by the respective kernel.
    The antenna aperture illumination kernels do not have to be identical to each
    other. A discrete set of arbitrarily placed antennas are now placed on to a
    regular grid.}
  \label{fig:a2g-mapping}
\end{figure}

\par\medskip
\noindent {\bf Temporal Fourier transform}
\par\medskip
\noindent This module is common to the MOFF and visibility-based imaging 
techniques. Time samples of electric fields measured by the antenna and 
digitized by the A/D converter is Fourier transformed to generate electric 
field spectra. This step can be parallelized across antennas as shown in 
Fig.~\ref{fig:f-engine}. The output is then fed to either MOFF or 
visibility-based imaging pipelines.

\begin{figure}
  \includegraphics[width=\columnwidth]{figure4}
  % \includegraphics[width=\columnwidth]{F-engine}
  % \includegraphics[width=\columnwidth]{F-engine.eps}
  \caption{Block diagram of a F-engine. The electric field data streams from
    antennas are Fourier transformed in parallel to generate electric field
    spectra.}
  \label{fig:f-engine}
\end{figure}

@@

In a companion paper to appear soon, we demonstrate a novel calibration 
technique (EPICal) which leverages the data products formed by direct imaging 
correlators to estimate antenna complex gains. This method correlates the antenna 
electric field signals with an image pixel from the output of the correlator in 
the feedback calibration fashion outlined in \citealt{mor11} (illustrated in 
Fig.~\ref{fig:MOFF-flowchart} by the arrow leading from the imager to the 
calibration block). Furthermore it allows for arbitrarily complex sky models, 
and following the MOFF algorithm places no restriction on array layout, and 
accounts for non-identical antenna beam patterns. Direction dependent calibration 
can be achieved by correlating antenna signals with output pixels in the 
direction of $N_\textrm{c}$ calibration sources, then fitting for a functional 
model of the sky. Since antennas are only correlated with calibrator pixels, the 
computational complexity scales as $\sim N_\textrm{a} N_\textrm{c}$.

@@

In MOFF imaging, these are spatially Fourier transformed followed by a squaring
operation at every time stamp for every frequency channel. In visibility-based 
imaging, the spatial Fourier transform is performed only once per integration 
time-scale and does not include a squaring operation.

@@

In contrast, visibilities measured by an antenna are statistically stable within
an integration time interval. Hence, they are averaged after calibration as shown
in equation~\ref{eqn:cc-vis}. It is advantageous to average them in visibilities
before imaging because the repeated cost of spatial FFT can be avoided. Since 
this averaging has been performed already on the visibilities over an integration 
time-scale, the imaging step has to be performed only once per integration cycle.

@@

In order to verify the accuracy of the EPIC code, we characterize the images 
produced through simulations. We simulate electric field streams from a model 
sky and process the data through both the MOFF and a visibility-based imaging 
algorithm. We then compare the output images to demonstrate their equivalence.

@@

We use the EPIC simulator to generate stochastic electric field samples from a 
sky model consisting of 10 point sources of flux densities 10~Jy each at random 
locations. In our simulations, we use 64 frequency channels each of width 
$\Delta f = 40$~kHz. The number of time stamps integrated in one integration 
cycle was kept at eight where each A/D timeseries is $1/\Delta f=25\,\mu$s long. 
We use the MWA array layout \citep{bea12} for demonstration. Only the inner 51 
tiles within a square bounding box of 150~m on each side were used. We assumed 
all tiles are identical and have a square shaped electric field illumination 
footprint 4.4~m on each side. Besides the stochastic sky noise present in the 
simulated electric fields, no noise from the instrument is added.

\subsection{Antenna auto-correlations}\label{sec:rm-autocorr}

Before the outputs can be compared, we describe the elimination of a minor 
mathematical difference between the two techniques. The squaring operation under 
MOFF imaging in the image plane introduces antenna auto-correlations around the 
zero spacing in the $uv$-plane which are absent in traditional visibility-based 
imaging. In order to facilitate a robust comparison between MOFF and 
visibility-based imaging techniques, these auto-correlations are removed from 
the MOFF algorithm output, which is otherwise not an essential part of the core 
algorithm. We describe below how they are removed.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figure5}
  % \includegraphics[width=\columnwidth]{autocorr_uvwts_pbeam}
  % \includegraphics[width=\columnwidth]{autocorr_uvwts_pbeam.png}
  \caption{The auto-correlation of weights of a square shaped antenna aperture
    in the $uv$ plane (left) and the corresponding directional antenna power 
    response on the sky (right) in coordinates specified by direction cosines. 
    The antenna auto-correlation weights are normalized to a sum
    of unity yielding a peak response of unity in the antenna's directional
    power pattern on the sky. The color scale for the directional power 
    pattern is logarithmic. The black circle indicates the sky horizon and
    values beyond it are not physical and hence ignored.}
  \label{fig:autocorr_wts_PB}
\end{figure}

We inverse Fourier transform the squared images and beams back to the $uv$ 
plane and subtract the estimated auto-correlation kernel scaled to the peak 
value centreed at the zero spacing pixel. The final averaged image is obtained 
by Fourier transforming the $uv$ plane data and weights with the 
auto-correlations subtracted to the image plane. These images are now comparable 
to those obtained from visibility-based imaging. This step of removing 
auto-correlations is optional and required to be performed only once per 
integration time-scale and does not add significant cost to the full operation.

@@

Fig.~\ref{fig:MOFF-FX-image} shows the dirty images (top) and synthesized beams
(bottom) obtained with antenna-based MOFF and FX visibility-based imaging 
algorithms packaged in EPIC. The antenna auto-correlations that correspond to 
zero spacing have been removed from the correlated weights and data in the 
$uv$ plane, the MOFF image and the corresponding synthesized beam as 
described in \S\ref{sec:rm-autocorr}. The sky positions of the simulated sources 
are indicated by solid black circles. The reconstructed sky image has the 
simulated sources at the expected sky positions in either case. Both algorithms 
result in images and synthesized beams that are well matched with each other. 
Their fluxes are modulated by a multiplicative power pattern corresponding to 
that of a uniform square aperture.

\begin{figure}
  \includegraphics[width=\columnwidth]{figure6}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_10_random_source_positions_4_iterations_test_aperture_zoomed}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_10_random_source_positions_4_iterations_test_aperture_zoomed.eps}
  \caption{Dirty images (top) and synthesized beams (bottom) obtained from 
    simulated data using EPIC implementation of antenna-based MOFF algorithm 
    (left) and visibility-based imaging (right). The solid black circles in the
    top panels indicate the simulated source positions. The antenna 
    auto-correlations at zero-spacing have been removed from the MOFF images. 
    The images in either case reconstruct the sources at the right locations 
    with the fluxes expected after multiplication by the antenna power pattern. 
    The synthesized beams from the two algorithms are well matched in size and
    shape. The overall modulation by the power pattern is seen clearly in both
    images.}
  \label{fig:MOFF-FX-image}
\end{figure}

Now we investigate the gridded cross-correlation weights in the $uv$-plane. 
Fig.~\ref{fig:MOFF-FX-uvwts} shows the correlated weights obtained with MOFF 
imaging (left) and visibility-based imaging (right). There are two notable 
differences. The first is in the weights around zero spacing. Though both show a
dominant void around zero-spacing, the void obtained with MOFF algorithm shows 
many pixels with non-zero weights. In contrast, the zero-spacing void from 
traditional imaging consists of predominantly zero-valued pixels. The second 
difference is in the weights outside the zero-spacing void. In some regions they 
are different from each other at the few percent level though the sum of weights 
in these regions is the same. For instance, note the difference in weights at 
$(u,v)\approx (2,-10)$. The reason for these differences is described below.

The gridding step in MOFF imaging samples the antenna footprint (either in 
analytic or lookup table formats) at the grid locations. Coverage of grid pixels 
by an antenna footprint may be $\sim 1$ pixel narrower particularly at the 
edge of the footprint along one or both directions relative to that from another 
identical antenna but with a fractionally different location relative to the 
grid. This depends on the exact location of the centre of the antenna relative 
to the grid and the coarseness of grid spacing. This first order loss of 
precision of the sampled footprint propagates to second order ($\sim 2$ pixels) 
upon correlation of the discretized weights. In other words, the correlated 
weights may suffer further loss of precision in their sampled footprint after 
correlation of two footprints each of which could be less precise to first 
order. On the other hand, in visibility-based imaging, a directly sampled $uv$ 
plane antenna power response (which is theoretically identical to the correlation 
of individual antenna footprints) centreed on a baseline has a loss in precision 
at most to first order. Thus, although in the limit of infinitesimally small 
grid spacing they should be identical, the coarseness of grid spacing introduces 
subtle differences between the two.

These differences which are dependent on the coarseness of grid spacing can be 
mitigated by making the grid spacing finer at the expense of increased 
computational cost. Residuals centreed around zero spacing can also be lowered 
by subtracting each auto-correlation of antenna weights separately by using the 
shape and extent of the sampled footprint appropriate for that specific antenna 
aperture. This is a general solution applicable even in case of heterogeneous 
antenna arrays and is under active development for EPIC.
